### 2016-09-22 doc/publications/plos16/plos16.tex
small changes to plos16.tex

Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to prioritise SNPs for their contribution to disease. 
The joint interaction is achieved by appling the technique of bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example, by summing the number of times a variable is used to split a node, over all nodes and all trees.
RF are well suited for processing ``wide'' genomic data for two reasons. 
Firstly, while other machine learning applications have the propensity to overfit datasets with more features $p$ than samples $n$ (``curse of dimensionality''~\cite{Bauer2014, bellman1961adaptive}), RF are known to be more resistant through not immune to overfitting~\cite{Segal.2004}.
Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow separate trees on each processor and combine the mature trees.

The most popular random forest software packages are R-based, although current implementations do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}. 
This is despite R now supporting long vectors (length greater than $2^{31}$), and the standard {\sc random forest} implementation in R using paralelization to grow multiple trees simultaneously~\cite{Liaw2002}. 
%While R has provisions to support long vectors (length greater than $2^{31}$), current R-based RF implementations, such as {\sc random forest} or {\sc party}~\cite{Hothorn2006}, do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.
More suitable implementations for ``wide'' genomic data are hence developed in other programming languages.  
{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by these methods. 
The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) as oppose to using virtually infinite numbers of CPUs across multiple nodes. This in turn limits the number of tasks that can be performed in parallel such that only splitting at whole tree-level is feasible as oppose to a more fine grained strategy of handing off the calculation for each tree node to a separate processor. 
Hadoop/Spark overcomes these limitation by enabling programs to scale beyond compute-node boundaries and hence enable more sophisticated parallelisation strategies.  
%A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for ``big" (many samples) and ``wide" (many features) data sets. \cursedforest\ is capable of parallelising the split for each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}. 
By also utilising Spark to read in and manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small datasets where multi-threading is considered more performant. 
\marginpar{Piotr this is your 20 trees has the important feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be a result section.}
%Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller datasets. 
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees, which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis, VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning. 
This offers a comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be utilised to ``fill in" or impute the information at previously unobserved genomic positions in older array technology~\cite{Howie2012}. 
This provides the potential to impute samples in the GWAs catalogue and generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.

%(80000000 feat * 26.423 s)/167698 feat =12605 s = 3.5 hours
{\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.%, it processes 167,698 features in 26.423s.  
%DENIS Rob do you want to flesh this out more
\marginpar{Rob do you want to flesh this out more}
Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for dealing with large data sets.

However, these parallelisation strategies are capable to only utilise the CPUs on one compute node (multi-threading), which we show provides faster compute time on smaller dataset but becomes the bottleneck for datasets comprising of millions of features. 
Hadoop/Spark overcomes this limitation by enabling programs to scale beyond compute-node boundaries and virtually utilise infinite number of CPUs.   
A Spark application runs on a ``driver'' node, and this node divides and distributes work out to the many ``worker'' nodes, or ``executors''.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Method
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}

>>>>

\marginpar{decison trees capture interaction -- there is no main effect term. RF inherit this property. I dont think it
  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
Random Forests~\cite{Breiman2001} (RF) are able to capture interactions between features while still allowing to
prioritise SNPs for their contribution to disease.  The joint interaction is achieved by appling the technique of
bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
by summing the number of times a variable is used to split a node, over all nodes and all trees.  RF are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of dimensionality''~\cite{Bauer2014,
  bellman1961adaptive}), decision trees are resistant to overfitting. Random Forests are not immune to
overfitting as shown by ~\cite{Segal.2004} who gives  examples of the case where:
\begin{itemize}
\item a single tree is the most appropriate model; and
\item the varioable are hightly correlated that the boostrap resampling does not give rise to uncorrelated trees.
\end{itemize}
However, keeping these difficult cases in mind, RF have a strong calim on being resistant to overfitting.
Secondly, RF are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
separate trees on each processor and combine the mature trees.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}$), and the {\sc randomForest} implementation in R lending itself using paralelization to grow multiple trees
simultaneously~\cite{Liaw2002} and them combine them.
More suitable implementations for ``wide'' genomic data are hence developed in other programming languages.  {\sc
  CloudForest}~\cite{Bressler2015} written in Go, achieves fast running times by effective use of the CPU cache,
optimising for different classes of features and efficiently multi-threading.
Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016}
written in C++ with an R front-end is specifically optimised for dealing with large data sets.

However, the use of traditional compute infrastructure limits the parallelisation strategies that can be employed by
these methods.  The programs are limited to utilising only CPUs that are on the same computer node (multi-threading) or
else they run on CPUs distributed across nodes by virtue of the fact that there is no communication between the processes (a
separate tree is grown on each node).  Hadoop/Spark overcomes these limitation by enabling programs to scale beyond
compute-node boundaries and hence enable more sophisticated parallelisation strategies.
In the case of Random forests this measn that the the comutations for each node of a tree can be handed off to separate processors.

Here we introduce \cursedforest, a Hadoop/Spark based implementation of random forest specifically designed to cater for
``big'' (many samples) and ``wide'' (many features) data sets. \cursedforest\ is capable of parallelising the split for
each node in a tree thereby being able to handle millions of features, such as in the 1000 Genome data, which consists
of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}.  By also utilising Spark to read in and
manipulate the standard genomic variant format (VCF) directly, \cursedforest\ outperforms existing tools on even small
datasets where multi-threading is considered more performant.  \marginpar{Piotr this is your 20 trees has the important
  feature in the top 5 and 100 trees has it always as the top element, do we need to spell this out better, it should be
  a result section.}
%Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
%which boosts the accuracy for variable selection as demonstrated by {\sc r2VIM}~\cite{Szymczak2016} on smaller
%datasets. 
Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest\ can build a large number of trees,
which boosts the accuracy for variable importance selection~\cite{Szymczak2016}.

\cursedforest\ is included in our previously developed framework for Spark-based genomic data analysis,
VariantSpark~\cite{OBrien2015}, which therefore now enables supervised and unsupervised machine learning.  This offers a
comprehensive analysis toolkit that can scale for the future data demands, where the richness from WGS data can be
utilised to ``fill in'' or impute the information at previously unobserved genomic positions in older array
technology~\cite{Howie2012}.  This provides the potential to impute samples in the GWAs catalogue and generate datasets
of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern
compute paradigms to deal with these challenges.

A Spark application runs on a ``driver'' node, and this node
divides and distributes work out to the many ``worker'' nodes, or ``executors''.

@@

\section{Methods}
